---
title: "Gov 50 Final Project"
author: Emily Arbuiso
description: "Factors Contributing to New York's Air Pollution"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

Climate change and its related problems are becoming increasingly important to understand. Air pollution is one of those problems, and poses extreme risks to human health, such as increasing the chances the chances of heart disease or respiratory issues. Air pollution can decrease biodiversity, destroy crops, and generally poses many risks to important biological functions that keep the earth running properly. Besides the obvious release of harmful particles from human activity that cause air pollution, I would like to explore different factors that relate to air pollution. 

How do important socioeconomic factors such as unemployment, median income, post-secondary education, poverty levels, and population density impact air pollution? Which of these factors have the closest relationship to air pollution levels? I will explore these questions in this project by analyzing data from the 62 counties in New York State. 

## Description of Data

My final dataset, called finalData, is composed of data taken from the [Office of the New York State Comptroller Website](https://www.osc.state.ny.us/local-government/data) and the [Environmental Protection Agency Website](https://www.epa.gov/), and displays the following data for each of the 62 counties in New York for the year 2021: 

1. Unemployment = The percentage of county that is unemployed.
2. MedianIncome = The median income for the entire county, in dollars.
3. CollegeDegree = The proportion of the population that holds a college-level degree or higher. 
4. PM_2.5 = The level of fine particles with diameters of less than 2.5 micrometers in parts per million. PM_2.5 is a common air pollutant. 
5. Poverty = The percentage of the population that is in poverty. 
6. PopDensity = The population density of the county, measured in persons per square mile. 

This data represents an observational study. This means that no treatment was given, and data was collected through observation. This data represents a cross sectional-research design because it is taking data from one point in time, the year 2021, for a defined population of New York State.

## Project thoughts

I am interested in exploring data related to air pollution. I am interested in seeing how air pollution levels vary across different counties in New York State. Finally, I am interested in what factors might determine this variation. The specific factors I will explore are population density, economic activity, poverty, and education levels.

## Proposal

How do different factors like population density, economic activity, poverty, and education level affect a county's air pollution in the state of New York? Which of these factors impact air pollution the most? I plan to examine how these different factors could impact the levels of certain air pollutants like NO2, 03, and PM2.5. I hypothesize that there will be negative correlations between air pollution and the factor of education level. Typically, with a higher educational background, one has more knowledge of the climate/environmental science and is more aware of the factors that impact air pollution. Additionally, those with higher educational backgrounds tend to vote for more pro-environmental people that will institute policies that will decrease air pollution. I hypothesize that there will be positive correlations between air pollution and the factors of population density, economic activity, and poverty. The explanatory variables of interest are population density, economic activity, poverty, education level. Population density will be measured by people per square mile. Economic activity will be measured by unemployment and median household income. Poverty will be measured by percent of people below the poverty line. Educational background will be measured by the percent of people who have completed college. The outcome variable of interest is air pollution. This will be measured by ppb (parts per billion) or ppm (parts per million) for different pollutants. The years for the data will likely be in the most recent 3 or 4 years. If I observe lower air pollution in counties with higher educational backgrounds, this would support my hypothesis. If I observe higher air pollution in countries with higher educational backgrounds, this would go against my hypothesis. If I observe higher air pollution in counties with higher population densities, more economic activity, and more poverty, this would support my hypothesis. If I observe lower air pollution in counties with lower values of these factors, this would go against my hypothesis. I will also explore which factors correlate the most with air pollution and which do not. 

## Loading data sets 

```{r}

## loading data 

library(tidyverse)

finalData <- read_csv("finalData.csv")

## getting rid of extra columns 

finalData <- finalData |>
  select(County:PopDensity)

```
## Visualization #1: Distribution of PM_2.5

```{r}

finalData |>
  ggplot(mapping = aes(x = PM)) +
  geom_histogram() +
  labs(x = "PM_2.5 Levels",
       y = "Frequency",
       title = "Frequency of PM_2.5 Levels Across NY State (2022)")

```

## Visualization #2: PM_2.5 vs. Population Density

```{r}

## population density prediction

finalData |>
  ggplot(aes(x = PopDensity,
             y = PM)) +
  geom_point(color = "steelblue1") +
  geom_smooth(method = "lm", se = FALSE, 
              color = "indianred1", size = 1.5) +
  labs(x = "Population Density",
       y = "PM_2.5 Levels",
       title = "PM_2.5 Levels vs. Population Density in NY Counties")

## population density regression

pop_reg <- lm(PM ~ PopDensity, data = finalData)
pop_reg

## correlation

cor(finalData$PopDensity, finalData$PM) 

```

## Results of Population Density Regression

There appears to be a correlation between Population Density and PM_2.5 levels. There is a positive relationship between the two variables, as indicated by the line of best fit on the graph above. Generally, as population density increases, PM_2.5 levels also increase. 

This positive correlation between the two variables is further supported by the linear regression, which indicated an increase in PM_2.5 levels by 4.932e-05 ppm for each 1 unit increase in population density. This number is small, but this was expected as there is a large range in population density. 

Finally, the correlation coefficient between the two variables was calculated to be about 0.520, which indicates a moderate association between the two variables. 

These findings support my predictions, as I predicted that there would be a positive correlation between the two variables. 
